# (PART) Getting Started {.unnumbered}

# Introduction {#intro}

## Motivation

*Why do we want a package of something that is already working well, such as PyTorch?*

There are several reasons, but the main one is to bring another machine learning framework to R. Probably it is just me but I feel _PyTorch_ very comfortable to work with. Feels pretty much like everything else in Python. Very Pythonic. I have tried other frameworks in R. The closest that matches a natural language like PyTorch, is [MXnet](https://mxnet.apache.org/versions/1.7.0/get_started?). Unfortunately, _MXnet_ it is the hardest to install and maintain after updates.

Yes. I could have worked directly with _PyTorch_ in a native Python environment, such as _Jupyter_ or _PyCharm_ or [vscode](https://code.visualstudio.com/docs/python/jupyter-support) notebooks but it very hard to quit **RMarkdown** once you get used to it. It is the real thing in regards to [literate programming](https://en.wikipedia.org/wiki/Literate_programming)and __reproducibility__. It does not only contribute to improving the quality of the code but establishes a workflow for a better understanding of a subject by your intended readers [@knuth1983], in what is been called the *literate programming paradigm* [@cordes1991].

This has the additional benefit of giving the ability to write combination of _Python_ and _R_ code together in the same document. There will times when it is better to create a class in _Python_; and other times where _R_ will be more convenient to handle a data structure. I show some examples using `data.frame` and `data.table` along with _PyTorch_ tensors.

## How do we start using `rTorch`

Start using `rTorch` is very simple. After installing the minimum system requirements -such as _conda_- you just call it with:


```{r call-rtorch, class.source="badCode"}
library(rTorch)
```

There are several way of testing if `rTorch` is up and running. Let's see some of them:

### Getting the PyTorch version

```{r}
rTorch::torch_version()
```

### PyTorch configuration

This will show the PyTorch version and the current version of Python installed, as well as the paths to folders where they reside.

```{r}
rTorch::torch_config()
```

____
## What can you do with `rTorch`

Practically, you can do everything you could with **PyTorch** within the **R** ecosystem. Additionally to the `rTorch` module, from where you can extract methods, functions and classes, there are available two more modules: `torchvision` and `np`, which is short for `numpy`. We could use the modules with:

```{r}
rTorch::torchvision
rTorch::np
rTorch::torch
```

We get a glimpse of the first lines of the `help("torch")` via a Python chunk:
```{python, help-python, engine="python3", output.lines=20:50}
help("torch")
```


```{r}
library(reticulate)
use_condaenv("r-torch", required = TRUE)

reticulate::py_discover_config()
reticulate::py_config()
```


```{python, engine="python3"}
a = 10
print(a)

import numpy as np
print(np.abs(-3.14))
```

```{python, engine="python3"}
import pandas as pd

x = pd.DataFrame()
print(x)
```


