# (PART) Getting Started {.unnumbered}

# Introduction {#intro}

*Last update: `r system("git log -1 --format=\"%ad (%h)\" -- 0101-intro.Rmd", intern = TRUE)`*

```{r output-lines, echo=FALSE}
# this is a knitr hook to extract selected lines from the Python help output
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

## Motivation

*Why do we want a package of something that is already working well, such as PyTorch?*

There are several reasons, but the main one is to bring another machine learning framework to R. Probably, it is just me but I feel *PyTorch* very comfortable to work with. Feels pretty much like everything else in Python. Very **pythonic**. I have tried other frameworks in R. The closest that matches a natural language like PyTorch, is [MXnet](https://mxnet.apache.org/versions/1.7.0/get_started?). Unfortunately, *MXnet* it is the hardest to install and maintain after updates.

Yes. I could have worked directly with *PyTorch* in a native Python environment, such as *Jupyter,* or *PyCharm,* or [vscode](https://code.visualstudio.com/docs/python/jupyter-support) notebooks but it very hard to quit **RMarkdown** once you get used to it. It is the real thing in regards to [literate programming](https://en.wikipedia.org/wiki/Literate_programming) and **reproducibility**. It does not only contribute to improving the quality of the code but establishes a workflow for a better understanding of a subject by your intended readers [@knuth1983], in what is been called the *literate programming paradigm* [@cordes1991].

This has the additional benefit of giving the ability to write combination of *Python* and *R* code together in the same document. There will be times when it is better to create a class in *Python*; and other times where *R* will be more convenient to handle a data structure. I show some examples using `data.frame` and `data.table` along with *PyTorch* tensors.

## Start using `rTorch`

Start using `rTorch` is very simple. After installing the minimum system requirements -such as *conda* -, you just call it with:

```{r call-rtorch, class.source="badCode"}
library(rTorch)
```

There are several ways of testing if `rTorch` is up and running. Let's see some of them:

### Get the PyTorch version

```{r}
rTorch::torch_version()
```

### PyTorch configuration

This will show the PyTorch version and the current version of Python installed, as well as the paths to folders where they reside.

```{r}
rTorch::torch_config()
```

------------------------------------------------------------------------

## What can you do with `rTorch`

Practically, you can do everything you could with **PyTorch** within the **R** ecosystem. Additionally to the `rTorch` module, from where you can extract methods, functions and classes, there are available two more modules: `torchvision` and `np`, which is short for `numpy`. We could use the modules with:

```{r}
rTorch::torchvision
rTorch::np
rTorch::torch
```

## Getting help

We get a glimpse of the first lines of the `help("torch")` via a Python chunk:

```{python, help-python, engine="python3", output.lines=c(3:10)}
help("torch")
```

```{python, engine="python3", output.lines=c(1:20)}
help("torch.tensor")
```

```{python, engine="python3", output.lines=c(1:20)}
help("torch.cat")
```

```{python, engine="python3", output.lines=c(1:25)}
help("numpy.arange")
```

Finally, these are the classes for the module `torchvision.datasets`. We are using Python to list them using the `help` function.

```{python, engine="python3", output.lines=c(1:35)}
help("torchvision.datasets")
```

In other words, all the functions, modules, classes in PyTorch are available to rTorch.
