# PyTorch and NumPy

## Callable PyTorch modules from rTorch

### The `torchvision` module

This is an example of using the `torchvision` module. With `torchvision` we could download any of the datasets made available by PyTorch. In this example, we will be downloading the training dataset of the **MNIST** handwritten digits. There are 60,000 images in the training set and 10,000 images in the test set. The images will download on the folder `./datasets`.

```{r download-mnist-train}
library(rTorch)

transforms  <- torchvision$transforms

# this is the folder where the datasets will be downloaded
local_folder <- './datasets/mnist_digits'

train_dataset = torchvision$datasets$MNIST(root = local_folder, 
                                           train = TRUE, 
                                           transform = transforms$ToTensor(),
                                           download = TRUE)

train_dataset
```

You can do similarly for the `test` dataset if you set the flag `train = FALSE`. The `test` dataset has only 10,000 images.

```{r download-mnist-test}
test_dataset = torchvision$datasets$MNIST(root = local_folder, 
                                          train = FALSE, 
                                          transform = transforms$ToTensor())
test_dataset
```

### The `numpy` module

`numpy` is automaticaly installed when `PyTorch` is. There is some interdependence between both. Anytime that we need to do some transformation that is not available in `PyTorch`, we will use `numpy`. Just keep in mind that `numpy` does not have support for *GPUs*.

There are several operations that we could perform with `numpy` such creating arrays:

#### Create an array {.unnumbered}
Create an array:

```{r}
# do some array manipulations with NumPy
a <- np$array(c(1:4))
a
```

Create an array of a desired shape:

```{r}
np$reshape(np$arange(0, 9), c(3L, 3L))
```

Create an array by spelling out its components and `type`:

```{r}
np$array(list(
             list(73, 67, 43),
             list(87, 134, 58),
             list(102, 43, 37),
             list(73, 67, 43), 
             list(91, 88, 64), 
             list(102, 43, 37), 
             list(69, 96, 70), 
             list(91, 88, 64), 
             list(102, 43, 37), 
             list(69, 96, 70)
           ), dtype='float32')
```

## Common array and tensor operations in NumPy and PyTorch

We will use the `train` and `test` datasets that we loaded with `torchvision`.

#### Reshape an array {.unnumbered}

For the same `test` dataset that we loaded above from `MNIST` digits, we will show the image of the handwritten digit and its label or class. Before plotting the image, we need to:

1.  Extract the image and label from the dataset
2.  Convert the tensor to a numpy array
3.  Reshape the tensor as a 2D array
4.  Plot the digit and its label


```{r fig.asp=1}
rotate <- function(x) t(apply(x, 2, rev))   # function to rotate the matrix

# label for the image
label <- test_dataset[0][[2]]
label    

# convert tensor to numpy array
.show_img <- test_dataset[0][[1]]$numpy()
dim(.show_img) 

# reshape 3D array to 2D 
show_img <- np$reshape(.show_img, c(28L, 28L))
dim(show_img)
```

```{r fig.asp=1}
# show in gray shades and rotate
image(rotate(show_img), col = gray.colors(64))
title(label)
```

#### Generate a random array in NumPy {.unnumbered}

```{r}
# set the seed
np$random$seed(123L)
# generate a random array
x = np$random$rand(100L)
x
# calculate the y array
y = np$sin(x) * np$power(x, 3L) + 3L * x + np$random$rand(100L) * 0.8
class(x)
class(y)
```

From the classes, we can tell that the `numpy` arrays are automatically converted to `R` arrays.

```{r, fig.asp=1}
plot(x, y)
```

#### Generate a random array in PyTorch {.unnumbered}

The same operation can be performed with pure `torch` tensors:

```{r, fig.asp=1, results="hold"}
library(rTorch)

invisible(torch$manual_seed(123L))
x <- torch$rand(100L)     # use torch$randn(100L): positive and negative numbers
y <- torch$sin(x) * torch$pow(x, 3L) + 3L * x + torch$rand(100L) * 0.8
class(x)
class(y)
```

Since the clasess are `torch` tensors, to plot them in R, they need to be converted to numpy, and then R:

```{r, fig.asp=1}
plot(x$numpy(), y$numpy())
```

#### Convert a `numpy` array to a PyTorch tensor {.unnumbered}

This is a very common operation that I have seen in examples using PyTorch. Creating first the array in `numpy`. and then convert it to a `torch` tensor.

```{r}
# input array
x = np$array(rbind(
            c(0,0,1),
            c(0,1,1),
            c(1,0,1),
            c(1,1,1)))

# the numpy array
x
```

This is another common operation that will find in the PyTorch tutorials: converting a `numpy` array from a certain type to a tensor of the same type:

```{r}
# convert the numpy array to a float type
Xn <- np$float32(x)
# convert the numpy array to a float tensor
Xt <- torch$FloatTensor(Xn)
Xt
```

## Python built-in functions

To access the Python built-in functions we make use of the package `reticulate` and the function `import_builtins()`.

Here are part of the built-in functions and operators offered by `reticulate`:

```{r}
py_bi <- import_builtins()
grep("Error|Warning|Exit", names(py_bi), value = TRUE, invert = TRUE, perl = TRUE)

```

#### Length of a dataset {.unnumbered}

Sometimes, we will need the Python `len` function to find out the length of an object:

```{r}
py_bi$len(train_dataset)
py_bi$len(test_dataset)
```

#### Iterators {.unnumbered}

Iterators are used a lot in dataset operations when running a neural network. In this example we will iterate through only 100 elements of the 60,000 of the `train` dataset. The goal is printing the "label" or "class" for the digits we are reading. The digits are not show here; they are stored in tensors.

<!-- we use results="hold" to wait for the header of the table -->

```{r, collapse=TRUE, results="hold"}
# iterate through training dataset
enum_train_dataset <- py_bi$enumerate(train_dataset)
cat(sprintf("%8s %8s \n", "index", "label"))

for (i in 1:py_bi$len(train_dataset)) {
    obj <- reticulate::iter_next(enum_train_dataset)
    idx   <- obj[[1]]        # index number
    cat(sprintf("%8d %5d \n", idx, obj[[2]][[2]]))
    if (i >= 100) break   # print only 100 labels
}
```

#### Types and instances {.unnumbered}

Types, instances and classes are important to take decisions on how we will process data that is being read from the datasets. In this example, we want to know if an object is of certain instance:

```{r}
# get the class of the object
py_bi$type(train_dataset)

# is train_dataset a torchvision dataset class
py_bi$isinstance(train_dataset, torchvision$datasets$mnist$MNIST)
```
