# Converting tensors

_Last update: `r system("git log -1 --format=\"%ad (%h)\" -- 0104-converting_tensors.Rmd", intern = TRUE)`_

```{r}
library(rTorch)
```


## Tensor to `numpy` array
This is a frequent operation. I have found that this is necessary when:

* a `numpy` function is not implemented in PyTorch
* We need to convert a tensor to R
* Perform a boolean operation that is not directly available in PyTorch

```{r}
x <- torch$arange(1, 10)
y <- x^2
```

If we attempt to plot these two tensors we get an error:

```{r, error=TRUE}
plot(x, y)
```
They need to be converted to `numpy`, and then to R (which happens in the background):

```{r}
plot(x$numpy(), y$numpy())
```

## `numpy` array to tensor
* Explain how transform a tensor back and forth to `numpy`.
* Why is this important?
* In what cases in this necessary?

```{r}
p <- np$arange(1, 10)
class(p)
```
```{r}
(pt <- torch$as_tensor(p))
```
```{r}
class(pt)
```


### `numpy` array to `R`
This is mainly required for these reasons:

1. Create a data structure in R
2. Plot using `r-base` or `ggplot2`
3. Perform an analysis on parts of a tensor
4. Use R statistical functions that are not available in PyTorch


## R objects to `numpy` objects

Given the R matrix $m$:

```{r}
m <- matrix(seq(1,10), nrow = 2)
m
```
We explicitly convert it to a `numpy` object with the function `r_to_py()`:

```{r}
(mp <- r_to_py(m))
```

```{r}
class(mp)
```

```{r}
class(mp)
```

