# Creating PyTorch classes

_Last update: `r system("git log -1 --format=\"%ad (%h)\" -- 0203-creating_pytorch_classes.Rmd", intern = TRUE)`_

## Build a PyTorch model class
PyTorch classes cannot not directly be instantiated from `R`. Yet. We need an intermediate step to create a class. For this, we use `reticulate` functions like `py_run_string()` that will read the class implementation in `Python` code, and then assign it to an R object.

### Example 1: One layer NN

```{r class-linreg, eval=FALSE}
py_run_string("import torch")
main = py_run_string(
"
import torch.nn as nn

class Net(nn.Module):
   def __init__(self):
       super(Net, self).__init__()
       self.layer = torch.nn.Linear(1, 1)

   def forward(self, x):
       x = self.layer(x)      
       return x
")


# build a Linear Rgression model
net <- main$Net()
```

The R object `net` now contains all the object in the PyTorch class `Net`.


### Example 2: Logistic Regression

```{r class-logreg, eval=FALSE}
main <- py_run_string(
"
import torch.nn as nn

class LogisticRegressionModel(nn.Module):
    def __init__(self, input_dim, output_dim):
        super(LogisticRegressionModel, self).__init__()
        self.linear = nn.Linear(input_dim, output_dim)

    def forward(self, x):
        out = self.linear(x)
        return out
")

# build a Logistic Rgression model
LogisticRegressionModel <- main$LogisticRegressionModel
```

The R object `LogisticRegressionModel` now contains all the objects in the PyTorch class `LogisticRegressionModel`.


