---
title: "A Minimal rTorch Book"
author: "Alfonso R. Reyes"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is a minimal tutorial about using the `rTorch` package to have fun while doing machine learning. This book was written with [bookdown]()."
---

```{r echo=FALSE}
reticulate::use_condaenv("r-torch", required = TRUE)
```

# Prerequisites {.unnumbered}

*Last update: `r system("git log -1 --format=\"%ad (%h)\" -- index.Rmd", intern = TRUE)`*

You need couple of things to get `rTorch` working:

1.  Install Python [Anaconda](https://www.anaconda.com/products/individual). Preferably, for 64-bits, and above Python 3.6+. I have successfully tested Anaconda under four different operating systems: Windows (Win10 and Windows Server 2008); macOS (Sierra, Mojave and Catalina); Linux (Debian, Fedora and Ubuntu); and lastly, Solaris 10. All these tests are required by CRAN.

2.  Install [R](), [Rtools]() and [RStudio](). I used two R versions R-3.6.3 and R-4.0.2.

3.  Install the R package [reticulate](https://github.com/rstudio/reticulate), which is the one that provides the connection between R and Python.

4.  Install the stable version `rTorch` from CRAN, or the latest version under development via GitHub.

> Note. While it is not mandatory to have a previously created a `Python` environment with `Anaconda`, where `PyTorch` and `TorchVision` have already been installed, it is another option if for some reason `reticulate` refuses to communicate with the conda environment. Keep in mind that you could also get the `rTorch` *conda* environment installed directly from the `R` console, in very similar fashion as in [R-TensorFlow]() does. Use the function `install_pytorch()` to install a conda environment for PyTorch.

## Installation {.unnumbered}

The **rTorch** package can be installed from CRAN or Github.

From CRAN:

```{r rtorch-install-cran, eval=FALSE}
install.packages("rTorch")
```

From GitHub, install `rTorch` with:

```{r rtorch-install-github, eval=FALSE}
devtools::install_github("f0nzie/rTorch")
```

which will install rTorch from the `main` or `master` branch.

If you want to play with the latest rTorch version, then install it from the `develop` branch, like this:

```{r install-github-develop, eval=FALSE}
devtools::install_github("f0nzie/rTorch", ref="develop")
```

or clone with Git from the terminal with:

```{bash github-clone, eval=FALSE}
git clone https://github.com/f0nzie/rTorch.git
```

This will allow you to build `rTorch` from source.

## Python Anaconda {.unnumbered}

If your preference is installing an Anaconda environment first, these are the steps:

### Example {.unnumbered}

1.  Create a `conda` environment from the terminal with:

```{bash, create-conda-env, eval=FALSE}
conda create -n r-torch python=3.7
```

2.  Activate the new environment with

```{bash, activate-conda-env, eval=FALSE}
conda activate r-torch
```

3.  Install the `PyTorch` related packages with:

```{bash, install-conda-packages, eval=FALSE}
conda install python=3.6.6 pytorch torchvision cpuonly matplotlib pandas -c pytorch
```

The last part `-c pytorch` specifies the **stable** *conda* channel to download the PyTorch packages. Your *conda* installation may not work if you don't indicate the channel.

Now, you can load `rTorch` in R or RStudio with:

```{r index-load-rtorch, eval=FALSE}
library(rTorch)
```

### Automatic installation {.unnumbered}

I used the idea for automatic installation in the `tensorflow` package for R, to create the function `rTorch::install_pytorch()`. This function will allow you to install a `conda` environment complete with all `PyTorch` requirements plus the packages you specify. Example:

```{r, eval=FALSE}
rTorch:::install_conda(package="pytorch=1.4", envname="r-torch", 
                       conda="auto", conda_python_version = "3.6", pip=FALSE, 
                       channel="pytorch", 
                       extra_packages=c("torchvision", 
                                        "cpuonly", 
                                        "matplotlib", 
                                        "pandas"))
```

This is explained in more detailed in the [rTorch package manual](https://f0nzie.github.io/rTorch/articles/installation.html).

> **Note.** `matplotlib` and `pandas` are not really necessary for `rTorch` to work, but I was asked if `matplotlib` or `pandas` could work with `PyTorch`. So, I decided to install them for testing and experimentation. They both work.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'rTorch','bookdown', 'knitr', 'rmarkdown'), 
  'packages.bib')
```
